Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-08 main.tex
Add more discussion to the results section.

The Axelrod library \cite{axelrodproject, knight2016open} is an open source
software for conducting iterated prisoner's dilemma (IPD) research with
reproducibility as a principal goal. Written in the Python programming language,
to date over the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

A total of \input{./assets/number_of_players.tex}are included, of which
\input{./assets/number_of_stochastic_players.tex}are stochastic.  In
Section~\ref{sec:standard} is concerned with the standard tournament with 200
turns whereas in Section~\ref{sec:noisy} a tournament with 5\% noise is
discussed. Due to the inherent stochasticity of these IPD tournaments, these
tournament were repeated \input{./assets/standard_number_of_repetitions.tex}
times. This allows for a detailed and confident analysis of the performance of
strategies. To illustrate the results considered,
Figure~\ref{fig:tit_for_tat_scores} shows the distribution of the mean score per
turn of Tit For Tat over all the repetitions. Similarly,
Figure~\ref{fig:tit_for_tat_ranks} shows the ranks of of Tit For Tat for each
repetition (we note that it never wins a tournament). Finally
Figure~\ref{fig:tit_for_tat_wins} shows the number of opponents beaten in any given
tournament: Tit For Tat does not win any match (this is due to the fact that it
will either draw with mutual cooperation or defect second).

@@

@@

% Table of best strategies
\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the trained strategies. Starred
        strategies $^{*}$
        indicates that the strategy was trained.}
        \label{tbl:standard_score}
\end{table}

Violin plots showing the distribution of the scores of each strategy (again
ranked by median score) are shown in Figure~\ref{fig:standard_boxplot}.

@@

Pairwise payoff results are given as a heatmap (Figure~\ref{fig:standard_heatmap})
which shows that many strategies achieve mutual cooperation. The top performing
strategies never defect first yet are able to exploit weaker strategies that
attempt to defect.

@@

The strategies that win the most matches are Defector, Aggravater, followed by
handshaking and zero determinant strategies. This includes two handshaking
strategies that were the result of training to maximize Moran process fixation.
No strategies were trained specifically to win matches. None of the top scoring
strategies appear in the top 20 list of strategies ranked by match wins.
This can be seen in Figure~\ref{fig:standard_winplot} where the distribution of
the number of wins of each strategy is shown.

@@

Table~\ref{tbl:standard_wins_top_winners} shows the same information as
Table~\ref{tbl:standard_wins} but for the top 15 strategies who win the most
head to head matches.
% TODO Add more discussion?

@@

Figure~\ref{fig:comparison_cooperation_heatmaps_standard} shows the rate of
cooperation in each round for the top three strategies.
% TODO Discuss more? (If we want to keep this)

@@

@@

We also ran noisy tournaments in which there is a 5\% chance that an action is
flipped. As shown in Table~\ref{tbl:noisy_score} and
Figure~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to correct for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also designed to handle noise).

@@

The strategies trained in the presence of noise are also among the best
performers in the absence of noise. As shown in Figure~\ref{fig:noisy_heatmap}
the cluster of mutually cooperative strategies is broken by the noise at 5\%. A
similar collection of players excels at winning matches but again they have a
poor total payoff.

@@

As shown in Figure~\ref{fig:noisy_winplot} the strategies tallying the most wins
are somewhat similar, with Defector, the handshaking CollectiveStrategy, and
Aggravate appearing as the top three again.

@@

Table~\ref{tbl:noisy_wins_top_winners} shows the same information as
Table~\ref{tbl:noisy_wins} but for the top 15 strategies who win the most
head to head matches.
% TODO Add more discussion?

@@

Again we compare the fingerprints. For the top performing noisy strategies
(Figure~\ref{fig:comparison_fingerprint_noisy}) there
is a striking similarity in the fingerprints which indicates that the strategies
may behave similarly in principle.

@@

Figure~\ref{fig:comparison_cooperation_heatmaps_noisy} shows the rate of
cooperation in each round for the top three strategies.
% TODO Add more discussion

>>>>

The iterated prisoner's dilemma (IPD) is an area of game theoretic research
aiming to understand the evolution of cooperative behaviour from complex
dynamics \cite{Axelrod1984}.

The Axelrod library \cite{axelrodproject, knight2016open} is an open source
software for conducting IPD research with
reproducibility as a principal goal. Written in the Python programming language,
to date over the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

A total of \input{./assets/number_of_players.tex}are included, of which
\input{./assets/number_of_stochastic_players.tex}are stochastic.  In
Section~\ref{sec:standard} is concerned with the standard tournament with 200
turns whereas in Section~\ref{sec:noise} a tournament with 5\% noise is
discussed. Due to the inherent stochasticity of these IPD tournaments, these
tournament were repeated \input{./assets/standard_number_of_repetitions.tex}
times. This allows for a detailed and confident analysis of the performance of
strategies. To illustrate the results considered,
Figure~\ref{fig:tit_for_tat_scores} shows the distribution of the mean score per
turn of Tit For Tat over all the repetitions. Similarly,
Figure~\ref{fig:tit_for_tat_ranks} shows the ranks of of Tit For Tat for each
repetition (we note that it never wins a tournament). Finally
Figure~\ref{fig:tit_for_tat_wins} shows the number of opponents beaten in any given
tournament: Tit For Tat does not win any match (this is due to the fact that it
will either draw with mutual cooperation or defect second).

@@

The utilities used are \((R, P, T, S)=(3, 1, 5, 0)\) thus the specific
Prisoner's Dilemma being played is:

\begin{equation}\label{equ:pd}
    \begin{pmatrix}
        (3, 3) & (0, 5)\\
        (5, 0) & (1, 1)
    \end{pmatrix}
\end{equation}

@@

% Table of best strategies
\begin{table}[!hbtp]
        \centering
        \input{./assets/standard_top_15_scores_summary.tex}
        \caption{Standard Tournament: Mean score per turn of top 15 strategies
            (ranked by median over
        \protect\input{./assets/standard_number_of_repetitions.tex}tournaments).
        The leaderboard is dominated by the trained strategies (indicated by a
        $^{*}$}
        \label{tbl:standard_score}
\end{table}

For completeness, violin plots showing the distribution of the scores of each
strategy (again ranked by median score) are shown in
Figure~\ref{fig:standard_boxplot}.

@@

Pairwise payoff results are given as a heatmap (Figure~\ref{fig:standard_heatmap})
which shows that many strategies achieve mutual cooperation (obtaining a score of 3). The top performing
strategies never defect first yet are able to exploit weaker strategies that
attempt to defect.

@@

The strategies that win the most matches
(Table~\ref{tbl:standard_wins_top_winners} are Defector and Aggravater, followed
by handshaking and zero determinant strategies. This includes two handshaking
strategies that were the result of training to maximize Moran process fixation
(TF1 and TF2). No strategies were trained specifically to win matches. None of
the top scoring strategies appear in the top 15 list of strategies ranked by
match wins. This can be seen in Figure~\ref{fig:standard_winplot} where the
distribution of the number of wins of each strategy is shown.

@@

@@

Figure~\ref{fig:comparison_cooperation_heatmaps_standard} shows the rate of
cooperation in each round for the top three strategies. The opponents in these
figures are ordered according to performance by median score. It is evident that
the high performing strategies share a common thread against the top strategies:
they do not defect first and achieve mutual cooperation. Against the lower
strategies they also do not defect first (a mean cooperation rate of 1 in the
first round) but do learn to quickly retaliate.

@@

% TODO Update these with 10k repetition version? (I have these for all but DBS)

@@

Noisy tournaments in which there is a 5\% chance that an action is
flipped are now described. As shown in Table~\ref{tbl:noisy_score} and
Figure~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to correct for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also designed to handle noise).

@@

Recalling Table~\ref{tbl:standard_score}, the strategies trained in the presence
of noise are also among the best performers in the absence of noise. As shown in
Figure~\ref{fig:noisy_heatmap} the cluster of mutually cooperative strategies is
broken by the noise at 5\%. A similar collection of players excels at winning
matches but again they have a poor total payoff.

@@

As shown in Table~\ref{tbl_noisy_wins_top_winners} and
Figure~\ref{fig:noisy_winplot} the strategies tallying the most wins are
somewhat similar, with Defector, the handshaking CollectiveStrategy, and
Aggravate appearing as the top three again.

@@

@@

Similarly to Section~\ref{sec:standard} the top performing noisy strategies
(Figure~\ref{fig:comparison_fingerprint_noisy}) have
a striking similarity in the fingerprints which indicates that the strategies
may behave similarly in principle.

@@

Figure~\ref{fig:comparison_cooperation_heatmaps_noisy} shows the rate of
cooperation in each round for the top three strategies and just as for the top
performing strategies in the standard tournament
(Figure~\ref{fig:comparison_cooperation_heatmaps_standard}) it is evident that
the strategies never defect first and learn to quickly punish poorer strategies.
